---
title: "Exploratory Data Analysis"
author: |
        | Felipe J Colon-Gonzalez
        | School of Environmental Sciences
date: "20-21 February 2017"
output: 
     ioslides_presentation:
          logo: uea_logo.png
---

# Part III. Exploratory Data Analysis

## Load a dataset
- Let's work with the `mtcars` set (*Motor Trend Car Road Tests*) from the 
`datasets` package
- Load the data and look at the dimensions and structure of the set
- Explore the description of the variables in the set

```r
> require(datasets)

> data(mtcars)

> ?mtcars
```

## Explore the set
- You can explore the top and bottom six rows using the `head()` and `tail()`
functions
- The number of rows can be modified with the argument `n=nrows` where nrows
is the number of rows you want to look at
- You can obtain information about each variable with `str()`

```r
 > str(mtcars)
'data.frame':	32 obs. of  11 variables:
 $ mpg : num  21 21 22.8 21.4 18.7 18.1 14.3 24.4 22.8 19.2 ...
 $ cyl : num  6 6 4 6 8 6 8 4 4 6 ...
 $ disp: num  160 160 108 258 360 ...
```


## Summary statistics
* Get summary statistics with the `summary()` function

```r
> summary(mtcars)
     mpg             cyl             disp             hp          
 Min.   :10.40   Min.   :4.000   Min.   : 71.1   Min.   : 52.0   
 1st Qu.:15.43   1st Qu.:4.000   1st Qu.:120.8   1st Qu.: 96.5  
 Median :19.20   Median :6.000   Median :196.3   Median :123.0  
 Mean   :20.09   Mean   :6.188   Mean   :230.7   Mean   :146.7  
 3rd Qu.:22.80   3rd Qu.:8.000   3rd Qu.:326.0   3rd Qu.:180.0  
 Max.   :33.90   Max.   :8.000   Max.   :472.0   Max.   :335.0  
 [... truncated]
 
```

## Summary statistics
* You can also get summary statistics independently using:
       - `mean()`
       - `sd()`
       - `var()`
       - `median()`
       - `range()`


## Frequency tables
* Create frequency tables with the `table()` function

```r
> table(mtcars$cyl)  # Number of cars by number of cylinders
 4  6  8 
11  7 14 

> with(mtcars, table(cyl, gear))  # Cars by cylinders and gears
   gear
cyl  3  4  5
  4  1  8  2
  6  2  4  1
  8 12  0  2
```


## Frequency tables
- 3 way contingency tables can be obtained with `ftable()`

```r
> ftable(mtcars[c("cyl", "am", "gear")])
       gear  3  4  5
cyl am              
4   0        1  2  0
    1        0  6  2
6   0        2  2  0
    1        0  2  1
8   0       12  0  0
    1        0  0  2
```


## Frequency tables
- The default `table()` call does not display information for missing data
- Frequencies for missing values can be obtained with `useNA="always"`

```r
> mtcars[1,2] <- NA
> table(mtcars$cyl)

 4  6  8 
11  6 14 

> table(mtcars$cyl, useNA="always")

   4    6    8 <NA> 
  11    6   14    1 
```


## Quantiles
- Sample quantiles for specific probabilities can be obtained with `quantile()` 

```r
> quantile(mtcars$mpg)        # Default
    0%    25%    50%    75%   100% 
10.400 15.425 19.200 22.800 33.900 
 
> quantile(mtcars$mpg, probs=c(0.1, 0.5, 0.9))
  10%   50%   90% 
14.34 19.20 30.09 
```
<br>
- **Now, try other quantiles**

## Evaluate logical vectors
- You may want to evaluate if a particular characteristic applies to some or 
all values in a vector

```r
> any(mtcars$mpg > 30)
[1] TRUE

> all(mtcars$mpg < 32)
[1] FALSE
```

# Exploratory plots

## Exploratory plots
- Simple exploratory plots can be easily generated using generic functions
such as `plot()`, `barplot()`, `hist()` and `boxplot()`
- Plots can be easily annotated and customized 
- One-dimension and two-dimensional plots may be generated


## Exploratory plots
- When no plot type is specified, R will use the plot type it thinks is most 
adequate for your data

```{r}
with(mtcars, plot(wt, mpg)) 
```

## Exploratory plots

```{r}
data(AirPassengers)
plot(AirPassengers)
```


## Customizing plots
- You may want your plot to show different colours based on a data feature.
- Using the `mtcars` dataset I indicate whether data correspond to automatic 
or manual cars

```{r, fig.height=3.5}
with(mtcars, plot(wt, mpg, pch=19, col=factor(am))) # ?pch
```

## Customizing plots
- You can also change the labels for the axes

```{r, fig.height=3.5}
with(mtcars, plot(wt, mpg, pch=19, col=factor(am), 
                  xlab="Weight", ylab="Miles per gallon"))
```


## Customizing plots
- Change the size of the points (equally or based on a condition)

```{r, fig.height=3.5}
with(mtcars, plot(wt, mpg, pch=19, col=factor(am), 
                  xlab="Weight", ylab="Miles per gallon",
                  cex=carb))
```

## Customizing plots
- Add a title

```{r, fig.height=3.5}
with(mtcars, plot(wt, mpg, pch=19, col=factor(am), 
                  xlab="Weight", ylab="Miles per gallon",
                  cex=carb, main="My first plot"))
```


## Customizing plots
- And a legend!

```{r, fig.height=3.5}
with(mtcars, plot(wt, mpg, pch=19, col=factor(am), 
                  xlab="Weight", ylab="Miles per gallon",
                  cex=carb, main="My first plot"))
legend("topright", legend=unique(mtcars$am), pch=19,
       col=c("black", "red"))
```


## Other plot types 
- One-dimensional box and whisker 

```{r}
data("ToothGrowth")
boxplot(ToothGrowth$len, col="blue")
```


## Two-dimensional box and whisker plots

```{r}
with(ToothGrowth, boxplot(len ~ supp, col=c("orange","blue")))
```


## Histogram

```{r}
x <- rnorm(1000)
hist(x)
```


## Histogram

```{r}
hist(x, breaks=50, col="green")
```


## Bar plots

```{r, fig.width=8,fig.height=5.5}
set.seed(234); x <- rnorm(100)
barplot(x, main="Random numbers")
```


## Scatter plot matrices

```{r}
pairs(mtcars[,3:6])
```


## Overlaying features
- Vertical line

```{r}
hist(x, breaks=15, col="steelblue4")
abline(v=median(x), col="tomato", lwd=6)
```


## Overlaying features
- Horizontal dotted line

```{r}
with(ToothGrowth, boxplot(len ~ supp, 
                          col=c("goldenrod","darkolivegreen")))
abline(h=28, col="red3", lwd=3, lty=3)
```


## Conditional colours

```{r, fig.width=8,fig.height=5.5}
set.seed(234); x <- rnorm(100)
barplot(x, main="Random numbers", col=ifelse(x<0, "blue", "red"))
```



## Multiple plots in one graphic device

```{r, fig.height=4, fig.width=8}
par(mfrow=c(1, 2)) # One device, one row, two columns
with(mtcars, {
       plot(wt, mpg, main="Weight vs. miles per gallon")
       plot(hp, mpg, main="Horsepower vs. miles per gallon")
})
```


## Graphic devices
- You may open more than one screen devices at once
       - On Mac the screen device is launched with `quartz()`
       - On Windows, the command is `windows()`
       - On Linux/Unix, the command is `x11()`  



## Save plots to a file

<br>
```r
> pdf("myplot.pdf")    # Opens graphics device in working directory

> with(mtcars, plot(wt, mpg, pch=19, col=cyl, cex=2))

> dev.off()            # Closes graphics device

# Open the plot on your computer
```
<br>
-  Now, explore other graphic devices and their options



## Your turn!
- Load the `mydata.csv` file
- Compute **summary statistics** for both sulfate and nitrate measurements (are
there any missing values?)
- Create a subset called `mySubset` with no missing observations 
- Compute a frequency table for each ID level on your subset
- Set the dates in the dataset as R-dates
- Generate a 2 panel plot with the histograms of sulfate and nitrate
- Create a subset where all negative sulfate values are equal to zero and 
plot sulfate values with colours by ID


## To practice at your own pace

- Install the interactive `swirl` package
- Go through the `R programming` modules


## Useful `R`esources
[Coursera](https://www.coursera.org/)

[DataCamp](https://www.datacamp.com/)

[swirl](http://swirlstats.com/)

[Institute for Digital Research and Education](http://www.ats.ucla.edu/stat/r/modules/)

[Try R Code School](http://tryr.codeschool.com/)


## References
1. [R Programming for Data Science](https://leanpub.com/rprogramming) Peng (2016)
2. [An introduction to R](https://cran.r-project.org/doc/manuals/r-release/R-intro.html) R Core Team (2016)
3. [A Beginner's Guide to R](http://www.springer.com/us/book/9780387938363) Zuur et al., (2009)

# End of Part III



